\subsection{Testing of I/O Module}
\subsubsection{Testbench Setup}
There is no special need to setup a testbench.
Thus, in this testing, the required data will be fed to the inputs of the module and the outputs will be monitored.
The following tests are performed. Special I/O SPI and I$^2$C testings are ommitted since they will require a seperate slave device.
Their indivitual testings have be performed in the previous sections
\begin{itemize}
    \item General I/O Output testings
    \item General I/O Input testings
    \item Special I/O UART testings
\end{itemize}

\newpage
\subsubsection{General I/O Output Testing}
In this test, the general I/O ports is set as output and required to produce the bit sequence of ``0000010101000101''.

\insertWaveform{general_io_output}{0.8}{Output Testing of General I/O}{waveform:general_io_output}

The results can be confirmed by examining the simulated waveform in Figure \ref{waveform:general_io_output}.
It can be noticed that the output is synchronous to clock.

\subsubsection{General I/O Input Testing}
In this test, the lower 5 bits of the general I/O is set as inputs, and there will be ``10101'' on the lower five pins.
Thus, the data that will be read from the input will be ``0000000000010101''.

\insertWaveform{general_io_input}{0.8}{Input Testing of General I/O}{waveform:general_io_input}
The results can be confirmed by examining the simulated waveform in Figure \ref{waveform:general_io_output}.
It can be noted that the input is asynchronous inside the I/O Module.

\subsubsection{Special I/O UART Transmission Testing}
In this test, a data of 230 will be loaded into UART data write register, and that data will be transmitted.
Thus, the UART transmitter will transmit, ``11100110'', ``00000000'', ``00000000'' and ``00000000'', since the total amount of data is 32 bits.

\insertWaveform{special_io_uart_tx}{0.8}{UART Transmission Testing of Special I/O}{waveform:special_io_uart_tx}
As can be observed in Figure \ref{waveform:special_io_uart_tx}, the transmission is correctly working.
Four pockets of data can be seen, each with the start and end bits. Little endianess is also displayed.
And the correct transmit bit pattern is also observed.