\newpage
\subsection{Testings of SPI Module}
Various tests will be performed on theSPI controller module. Unlike UART Module testings, we cannot do data loop back tests. A peripheral device needs to be built in order to simulate the overall operation. The implementation of this peripheral can be found on \ref{appendix:spi_peripheral_implementation}. The following tests will be performed.
\begin{enumerate}
	\item Clock Edge Generation
	\item Transmitting Data Once
	\item Transmitting Data Twice
	\item Reading Data Once
\end{enumerate}

\subsubsection{Testbench Setup}
The testbench is setup with SPI controller and SPI peripheral device. The necessary signals will be generated within the testbench.

\insertBlockDiagram{spi_module_tb}{0.6}{Block Diagram of testbench for SPI Module Testing}{block_diagram:spi_module_tb}

\subsubsection{Clock Edge Generation}
This test is to demonstrate the generation of the require clock edges and SPI clock for the module. The number of clock edges has to be 16, since there will be 8 bit data, so 8 clock cycles with 2 clock edges each. The counters can also be seen in figure \ref{waveform:spi_clock_edges}
\insertWaveform{spi_clock_edges}{1}{Testing Clock Edges for SPI Module}{waveform:spi_clock_edges}

\newpage
\subsubsection{Transmitting Data Once}
In this test, one data is sent to the peripheral device once. This test is to check the operation of write for one data.
\insertWaveform{spi_transmit_once}{1}{Transmitting Data Once}{waveform:spi_transmit_once}

\subsubsection{Transmitting Data Twice}
In this test, two pieces of data are transmitted consecutively to the same peripheral device. To choose the desired peripheral device, respective \io{ss\_n} pin is pulled low during transmission.

\insertWaveform{spi_transmit_twice}{1}{Transmitting Data Twice}{wave_form:spi_transmit_twice}

\newpage
\subsubsection{Reading Data Once}
In this test, a read is performed. Notable feature of SPI while reading is that it is not possible to read data without a write, due to simultaneous full-duplex nature of the protocol. Some form of data, either empty data or right data, need to be sent and the transmission and reception will take place at the same time on \io{mosi} and \io{miso} lines.

\insertWaveform{spi_read_once}{1}{Reading Data Once}{wave_form:spi_read_once}
